---
layout: post
title: "Calculating rotation invariant GLCM textures"
description: "Calculating rotation invariant GLCM textures in R"
category: articles
tags: [R, glcm, remote sensing]
comments: true
share: true
---

This post outlines how to use the `glcm` package to calculate image textures 
that are direction invariant (calculated over "all directions"). This feature 
is only available in `glcm` versions >= 1.0.

## Getting started

First use install the latest version of `glcm`, and the `raster` package that 
is also needed for this example:

```{r}
install.packages("glcm")

library(glcm)
library(raster)
```

## Calculating rotationally invariant textures

`glcm` supports calculating GLCMs using multiple shift values.  If multiple 
shifts are supplied, `glcm` will calculate each texture statistic using each of 
the specified shifts, and return the mean value of the texture for each pixel.  
In general, I have not found large differences in calculated image textures 
when comparing GLCM textures calculated using a single shift versus calculating 
rotationally invariant textures. However this may not be the case for images 
with strongly directional textures.

To compare for a sample cropped out of a Landsat scene, use the L5TSR_1986 
sample image included in the `glcm` package. This is a section of a 1986 
Landsat 5 image preprocessed to surface reflectance. The image is from the 
[Volcán Barva TEAM 
site](http://www.teamnetwork.org/network/sites/volc%C3%A1n-barva).

When `glcm` is run without specifing a shift, the default shift (1, 1) is used 
(90 degrees), with a window size of 3 pixels x 3 pixels. Below is an example 
from running `glcm` with the default parameters:

```{r, glcm_90deg_shift}
test_rast <- raster(L5TSR_1986, layer=1)
tex_shift1 <- glcm(test_rast)
plot(tex_shift1)
```

To calculate rotationally invariant GLCM textures (over "all directions" in the 
terminology of commonly used remote sensing software), use: `shift=list(c(0,1), 
c(1,1), c(1,0), c(1,-1))`. This will calculate the average GLCM texture using 
shifts of 0 degrees, 45 degrees, 90 degrees, and 135 degrees:

```{r, glcm_all_directions}
tex_all_dir <- glcm(test_rast, shift=list(c(0,1), c(1,1), c(1,0), c(1,-1)))
plot(tex_all_dir)
```

To compare the difference between these textures, subtract the textures 
calculated with a 90 degree shift from those calculated using multiple shifts, 
and plot the result:

```{r, glcm_all_directions_vs_90deg}
plot((tex_all_dir - tex_shift1) / tex_all_dir)
```

## Computation time

First look at the time difference for calculating a GLCM with only one shift 
versus calculating a rotationally invariant form:

```{r}
library(microbenchmark)

glcm_one_dir <- function(x) {
    glcm(x)
}

glcm_all_dir <- function(x) {
    glcm(x, shift=list(c(0,1), c(1,1), c(1,0), c(1,-1)))
}

microbenchmark(glcm_one_dir(test_rast), glcm_all_dir(test_rast), times=5)
```

As seen in the above, there is a performance penalty for using a rotationally 
invariant GLCM (not surprisingly, as more calculations are involved).

Prior to having the ability to use multiple shifts hardcoded in `glcm`, it was 
still possible to calculate rotationally invariant textures using the `glcm` 
function.  However, the calculation had to be done manually, using an approach 
similar to what I do below with `glcm_all_dir_manual`. How much faster is it 
perform the averaging directly in `glcm`?

```{r}
glcm_all_dir_manual <- function(x) {
    text_0deg <- glcm(x, shift=c(0,1))
    text_45deg <- glcm(x, shift=c(1,1))
    text_90deg <- glcm(x, shift=c(1,0))
    text_135deg <- glcm(x, shift=c(1,-1))
    overlay(text_0deg, text_45deg, text_90deg, text_135deg,
            fun=function(w, x, y, z) {
                return((w + x + y + z) / 4)
            })
}
tex_all_dir_manual <- glcm_all_dir_manual(test_rast)

# Check that the textures match
table(getValues(tex_all_dir_manual) == getValues(tex_all_dir))

microbenchmark(glcm_all_dir_manual(test_rast), glcm_all_dir(test_rast), 
               times=5)
```

The time difference isn't that great, but the need for repeated calls to `glcm` 
(and the need for multiple read/writes to disk for large files) could lead to a 
more substantial advantage for the direct approach with `glcm` than is apparent 
in this simple example. Of course, the manual approach does give more 
flexibility if you need to do other processing (or scaling, etc.) to the 
textures.
