---
layout: post
title: "Calculating image textures with GLCM"
description: "An overview of how to use the glcm R package to calculate image texture measures"
category: articles
tags: [R, teamlucc, remote sensing, glcm]
modified: 2014-03-19
comments: true
share: true
---

`glcm` can calculate image textures from either a matrix or a `Raster*` object 
from the `raster` package. First install the package if it is not yet 
installed:

```{r}
if (!(require(glcm))) install.packages("glcm")
```

The below examples use an image included in the `glcm` package, a 
red/green/blue cutout of a Landsat 5 image from 1986 from a Tropical Ecology 
Assessment and Monitoring (TEAM) Network site in Volcan Barva, Costa Rica. The 
image is included in the glcm package as `L5TSR_1986`:

```{r, L5TSR_1986_plot, fig.cap='1986 Landsat 5 image from Volcan Barva'}
library(raster) # needed for plotRGB function
plotRGB(L5TSR_1986, 3, 2, 1, stretch='lin')
```

To calculate GLCM textures from this image using the default settings, type:

```{r}
textures <- glcm(raster(L5TSR_1986, layer=3))
```

where `raster(L5TSR_1986, layer=3)` selects the third (red) layer.  To see the 
textures that have been calculated by default, type:

```{r}
names(textures)
```

This shows the eight GLCM texture statistics that have been calculated by 
default.  These can all be visualized in R:

```{r, mean, fig.cap="mean of GLCM texture"}
plot(textures$glcm_mean)
```

```{r, variance, fig.cap='variance of GLCM texture'}
plot(textures$glcm_variance)
```

```{r, homogeneity, fig.cap="homogeneity of GLCM texture"}
plot(textures$glcm_homogeneity)
```

```{r, contrast, fig.cap='contrast of GLCM texture'}
plot(textures$glcm_contrast)
```

```{r, dissimilarity, fig.cap='dissimilarity of GLCM texture'}
plot(textures$glcm_dissimilarity)
```

```{r, entropy, fig.cap='entropy of GLCM texture'}
plot(textures$glcm_entropy)
```

```{r, second_moment, fig.cap='second moment of GLCM texture'}
plot(textures$glcm_second_moment)
```

```{r, correlation, fig.cap='correlation of GLCM texture'}
plot(textures$glcm_correlation)
```
