---
layout: post
title: "Landsat Surface Reflectance CDR preprocessing with teamlucc"
description: "How to use the teamlucc to preprocess Landsat CDR surface reflectance imagery"
category: articles
tags: [R, teamlucc, remote sensing, Landsat]
comments: true
modified: 2014-11-13
share: true
---

## Overview

This post outlines how to use the `teamlucc` package to preprocess imagery from 
the [Landsat Surface Reflectance Climate Data Record 
(CDR)](http://landsat.usgs.gov/CDR_LSR.php) archive. The `teamlucc` package 
supports a number of common preprocessing steps, including:

* Conversion of CDR files to any GDAL supported file format
* Cropping Landsat tiles to a given area of interest (AOI)
* Mosaicking and cropping of DEM tiles (such as ASTER or SRTM) to a given AOI 
  or Landsat path/row
* Topographic correction of CDR scenes

## Getting started

First load the `teamlucc` package, and the `rgdal` package:

```{r}
library(teamlucc)
library(rgdal)
```

## DEM setup

Before CDR surface reflectance images can be topographically corrected with 
`teamlucc`, a digital elevation model (DEM) needs to be obtained that is in the 
same resolution and coordinate system as the CDR imagery, and slope and aspect 
need to be calculated. The `auto_setup_dem` function in the `teamlucc` package 
facilitates this task. There are many options you can supply to 
`auto_setup_dem` - see `?auto_setup_dem` for more information on these options. 
If you do not want to topographically correct your imagery, you can skip this 
step and move ahead to the next section.

`auto_setup_dem` assembles a DEM to cover a given area of interest (AOI), and 
can mosaic multiple DEM files if needed to cover an AOI. To use 
`auto_setup_dem`, the user must first define the area of interest with an AOI 
polygon. If you have a shapefile of an area of interest, load it into R using 
the `readOGR` command. The `readOGR` command needs the folder the shapefile is 
in (in this example the current working directory, specified by `.`) as the 
first parameter, and the filename of the shapefile (without the ".shp" 
extension) as the second parameter (`PA_VB`) in this example). This example 
uses a shapefile of the boundary of the Braulio Carrillo National Park, in 
which the [Volcán Barva TEAM 
site](http://www.teamnetwork.org/network/sites/volc%C3%A1n-barva) is located:

```{r, VB_aoi_original}
aoi <- readOGR('.', 'PA_VB')
plot(aoi)
```

As seen in the above plot, there are a number of adjoining polygons in this 
shapefile. The functions in `teamlucc` expect the AOI to be of length one. So 
calculate the convex hull of the AOI using the functions in the `rgeos` 
package:

```{r}
library(rgeos)
aoi <- gConvexHull(aoi)
```

The AOI is now a single polygon:

```{r, VB_aoi_convex_hull}
plot(aoi)
```

If you do not have an AOI, but know the Landsat path and row you want to work 
with, an alternative is to define the AOI based on the path and row, using the 
`wrspathrow` R package, and supplying the desired path and row numbers. Here 
`127` is the WRS-2 path number, and `47` is the WRS-2 row number for the 
path/row at the center of the above AOI:

```{r}
library(wrspathrow)
aoi_from_pathrow <- pathrow_poly(127, 47)
```

In addition to the AOI, `auto_setup_dem` needs to know the location and spatial 
extents of the DEM files you have available on your machine.  This list can be 
assembled automatically using the `get_extent_polys` function in `teamlucc`.  
See `?get_extent_polys` for more information. __The below code will fail on your 
machine because you will not have the proper DEMs for this example.  Download 
the proper DEMs for the area in which you are working and place them in a 
folder on your machine if you want to test this function.__

```{r}
dem_files <- dir('H:/Data/CGIAR_SRTM/Tiles', pattern='.tif$', full.names=TRUE)
dems <- lapply(dem_files, raster)
dem_extents <- get_extent_polys(dems)
```

For flexibility, `auto_setup_dem` can optionally crop each output DEM to the 
spatial extent of the supplied AOI. If `crop_to_aoi=TRUE`, then 
`auto_setup_dem` will crop the DEMs to the spatial extent of the supplied AOI. 
If `crop_to_aoi=FALSE`, then `auto_setup_dem` will crop the DEMs to the extent 
of the Landsat path/rows needed to cover the AOI.

Lastly, `auto_setup_dem` needs to know where to save its output. For this 
example, save the output to the current working directory `.`. Now that all of 
the essential inputs are defined, `auto_setup_dem` can be called:

```{r}
auto_setup_dem(aoi, '.', dem_extents, crop_to_aoi=TRUE)
```

The result will be a mosaicked DEM, in the current working directory, that can 
be used for topographically correcting imagery with the 
`auto_preprocess_landsat` function.

## Preprocessing

Images are delivered from the CDR archive in either ENVI format, GeoTIFF 
format, or Hierarchical Data Format (HDF). The `teamlucc` package will, by 
default, convert HDF or ENVI format images to a GeoTIFF format, as these image 
files can be easily read in R and in most commonly used remote sensing and GIS 
software packages. This example assumes you want GeoTIFF output - see the help 
files for `teamlucc` for other output options.

First you will need to acquire a time series of CDR imagery for your site. The 
`espa_download` function can facilitate downloading files from an ESPA order. 
Also see the post on [Filtering and downloading Landsat 
scenes](/articles/filtering-landsat-with-teamlucc) for more details on how 
`teamlucc` can help with image acquisition.

Once you have downloaded your imagery from ESPA, I recommend you put all of the 
zip files from your download in a single folder. You can then use the 
`teamlucc` `espa_extract` function to automate extracting your CDR image files, 
including placing the extracted files in consistently named output folders.  
If, for example, your files are located in `espa_downloads` and you want to 
extract them to `espa_extracts`, run the following (__this block, and the next 
block of code, will both fail on your computer since you do not have the 
required imagery - this is only an example, download your own imagery to follow 
along__):

```{r}
download_folder <- 'espa_downloads'
extract_folder <- 'espa_extracts'
espa_extract(download_folder, extract_folder)
```

Now that the CDR format image files are extracted, you are ready to run 
`auto_preprocess_landsat`. As with `auto_setup_dem`, there are many options you 
can supply to `auto_preprocess_landsat` - see `?auto_preprocess_landsat`.  The 
below is a simple example of how to call `auto_preprocess_landsat`.

The `image_dirs` line below is just a fancy way of finding all the folders 
located in `extract_folder` that contain CDR imagery.  You could just as easily 
specify these folders individually as as a list of strings, like: `image_dirs 
<- c('C:/folder1', 'C:/folder2')` if you had two CDR Landsat scenes located in 
`C:/folder1` and `C:/folder2`, respectively.

The `prefix` parameter specifies a string that will be used in naming files 
output by `auto_preprocess_landsat`. For `prefix` I suggest you use a short (2 
or 3 character) site name or site code that is meaningful to you.

There are two other options we provide below to `auto_preprocess_landsat`.  
`tc=TRUE` tells `auto_preprocess_landsat` to perform topographic correction.  
Because of this, we also need to specify `dem_path` (where the DEM files 
preprocessed by `auto_setup_dem` are located), so that the DEM files for this 
scene can be found. Here we set `dem_path='.'` as the DEM is in our current 
working directory. We supply an AOI (same AOI as above) to crop the output 
images. `verbose=TRUE` indicates that we want detailed progress messages to 
print while the script is running. The output of `auto_preprocess_landsat` is a 
`data.frame` with a list of the preprocessed files and their file formats.

```{r}
image_dirs <- dir('espa_extracts',
                  pattern='^[0-9]{3}-[0-9]{3}_[0-9]{4}-[0-9]{3}_((LT[45])|(LE7))$',
                  full.names=TRUE)
filelist <- auto_preprocess_landsat(image_dirs, prefix='VB', tc=TRUE, 
                                    dem_path='.', aoi=aoi, verbose=TRUE)
```

The result is two cropped, reprojected, and topographically corrected Landsat 
images covering the specified AOI. One image from 2000:

```{r, plot_2000_landsat}
ls_2000 <-  brick('espa_extracts/015-053_2000-044_LT5/VB_015-053_2000-044_L5TSR_tc.tif')
ls_2000 <- linear_stretch(ls_2000, pct=2, max_val=255)
browse_image(ls_2000, r=4, g=3, b=2)
```

And one image from 2001:

```{r, plot_2001_landsat}
ls_2001 <- brick('espa_extracts/015-053_2001-014_LT5/VB_015-053_2001-014_L5TSR_tc.tif')
ls_2001 <- linear_stretch(ls_2001, pct=2, max_val=255)
browse_image(ls_2001, r=4, g=3, b=2)
```

There is a fair amount of cloud cover in the 2000 image. See the post on [cloud 
removal](/articles/cloud-removal-with-teamlucc) for one means of addressing 
this issue.
