---
layout: post
title: "Analyzing forest change with gfcanalysis"
description: "How to use the gfcanalysis R package to analyze Hansen et al. 2013 Global Forest Change dataset"
category: articles
tags: [R, gfcanalysis, remote sensing, Landsat]
comments: true
share: true
---

## Overview
This `gfcanalysis` R package facilitates simple analyses using the Hansen et 
al. 2013[^1] [Global Forest Change 
dataset](http://earthenginepartners.appspot.com/science-2013-global-forest). 
The package was written to analyze forest change in within the Zone of 
Interaction surrounding each of the forest monitoring sites of the [Tropical 
Ecology Assessment and Monitoring (TEAM) Network](http://www.teamnetwork.org).

[^1]:
    Hansen, M. C., P. V. Potapov, R. Moore, M. Hancher, S. A. Turubanova, A. 
    Tyukavina, D. Thau, S. V. Stehman, S. J. Goetz, T. R. Loveland, A. Kommareddy, 
    A. Egorov, L. Chini, C. O. Justice, and J. R. G. Townshend. 2013. 
    High-Resolution Global Maps of 21st-Century Forest Cover Change. Science 342, 
    (15 November): 850--853. Data available on-line from: 
    http://earthenginepartners.appspot.com/science-2013-global-forest.

If you need help with any of the functions in the package, see the help files 
for more information. For example, type `?download_tiles` in R to see the help 
file for the `download_tiles` function.

## Getting started

This post will outline an analysis using the `gfcanalysis` package. Note that 
as the computations are intensive, some parts of this analysis may take some 
time to run (about 30 minutes total to run all of the code outlined here). If 
you do not already have the GFC product data downloaded on your computer, 
downloading the dataset will also take some time (though this process is 
automated by `gfcanalysis`).

To get started, first load the `devtools` package, used for installing 
`gfcanalysis`.  Install the `devtools` package if it is not already installed:

```{r}
if (!require(devtools)) install.packages('devtools')
```

Now load the gfcanalysis package, using `devtools` to install it from github if it 
is not yet installed. Also load the `rgdal` package needed for reading/writing 
shapefiles, and the `spatial.tools` package use for processing raster data in 
parallel:

```{r}
if (!require(gfcanalysis)) install_github('azvoleff/gfcanalysis')
if (!require(rgdal)) install.packages('rgdal')
if (!require(spatial.tools)) install.packages('spatial.tools')
```

Indicate where we want to save GFC tiles downloaded from Google. For any given 
AOI, the script will first check to see if these tiles are available locally 
(in the below folder) before downloading them from the server - so I recommend 
storing ALL of your GFC tiles in the same folder. For this example we will use 
"." - the current working directory of the R session.

```{r}
output_folder <- "."
```

Set the threshold for forest/non-forest based on the treecover2000 layer in 
the GFC product:

```{r}
forest_threshold <- 90
```

## Downloading data from Google server for a given AOI

Load an area of interest. For this example we use a shapefile of the Zone of 
Interaction (ZOI) of the [TEAM Network](http://www.teamnetwork.org) site in 
[Nam Kading National Protected Area, 
Laos](http://www.teamnetwork.org/network/sites/nam-kading-0). Notice that first 
we specify the folder the shapefile is in (here it is a "." indicating the 
current working directory), and then the name of the shapefile without the 
".shp". To follow along with this example, [download this shapefile of the 
ZOI](/content/2014-03-25-analyzing-forest-change-with-gfcanalysis/ZOI_NAK_2012_EEsimple.zip).

```{r, cache=TRUE}
aoi <- readOGR('.', 'ZOI_NAK_2012_EEsimple')
```

Calculate the tiles needed to cover the AOI:

```{r, cache=TRUE}
tiles <- calc_gfc_tiles(aoi)
print(length(tiles)) # Number of tiles needed to cover AOI
```

To check the overlap between the tiles and the aoi, you can make a plot of the 
needed tiles and the AOI using R's plotting functions:

```{r, tiles_versus_aoi, cache=TRUE}
plot(tiles)
plot(aoi, add=TRUE, lty=2, col="#00ff0050")
```

Now, check to see if these tiles are already present locally, and download them 
if they are not. By default the "first" and "last" composite surface 
reflectance images are not downloaded. To also download these images specify
`first_and_last=TRUE`.

```{r, cache=TRUE}
download_tiles(tiles, output_folder, first_and_last=FALSE)
```

## Performing thresholding and calculating basic statistics

Extract the GFC data for this AOI from the downloaded GFC tiles, mosaicing 
multiple tiles as necessary (if needed to cover the AOI). Save this extract in 
ENVI format in the current working directory (can also save as GeoTIFF, Erdas 
format, etc.)

```{r, cache=TRUE, cache.lazy=FALSE}
gfc_extract <- extract_gfc(aoi, output_folder, filename="NAK_GFC_extract.envi")
```

The extracted dataset has 5 layers (not yet thresholded):

```{r, cache=TRUE}
gfc_extract
```


Threshold the GFC data based on a specified percent cover threshold (0-100), 
and save the thresholded layers to an ENVI format raster:

```{r, cache=TRUE, cache.lazy=FALSE}
gfc_thresholded <- threshold_gfc(gfc_extract, forest_threshold=forest_threshold, 
                                 filename="NAK_GFC_extract_thresholded.envi")
```

```{r, echo=FALSE}
# Delete the gfc_extract ENVI files - don't need them
unlink("NAK_GFC_extract.*")
```

## Coding of the thresholded output

The thresholded dataset has 5 layers:

```{r, cache=TRUE}
gfc_thresholded
```

The output is coded using the following coding scheme:

### Band 1 (forest2000)

Based on the provided `forest_threshold`. Pixels wiwth percent canopy cover 
greater than `forest_threshold` are coded as forest.

| Cover in 2000         | Code |
|:----------------------|:----:|
| Non-forest            | 0 |
| Forest                | 1 |

### Band 2 (lossyear)

Note that lossyear is zero for pixels that were not forested in 2000.

| Year of loss  | Code |
|:--------------|:----:|
| No loss       |  0 |
| Loss in 2001  |  1 |
| Loss in 2002  |  2 |
| Loss in 2003  |  3 |
| Loss in 2004  |  4 |
| Loss in 2005  |  5 |
| Loss in 2006  |  6 |
| Loss in 2007  |  7 |
| Loss in 2008  |  8 |
| Loss in 2009  |  9 |
| Loss in 2010  | 10 |
| Loss in 2011  | 11 |
| Loss in 2012  | 12 |

### Band 3 (gain)

Note that gain is zero for pixels that were forested in 2000.

| Change    | Code |
|:----------|:----:|
| No gain  | 0 |
| Gain     | 1 |

### Band 4 (lossgain)

Note that loss and gain is difficult to interpret from the thresholded 
product, as the original GFC product does not provide information on the 
sequence of loss and gain (loss then gain, or gain then loss). The product also 
does not provide information on the levels of canopy cover reached prior to 
loss (for gain then loss) or after loss (for loss then gain pixels). The layer 
is calculated here as: `lossgain <- gain & (lossyear != 0)`, where `lossyear` 
and `gain` are the original GFC gain and lossyear layers, respectively.

| Change            | Code |
|:------------------|:----:|
| No loss and gain  | 0 |
| Loss and gain     | 1 |

### Band 5 (datamask)

| Class    | Code |
|:----------|:----:|
| No data  | 0 |
| Land     | 1 |
| Water    | 2 |


## Calculating statistics on forest loss and forest gain

Calculate annual statistics on forest loss/gain:
```{r, cache=TRUE}
gfc_stats <- gfc_stats(aoi, gfc_thresholded)
gfc_stats
```

Save these statistics to CSV files for use in Excel, or other software:

```{r, cache=TRUE}
write.csv(gfc_stats$loss_table, 
          file='NAK_GFC_extract_thresholded_losstable.csv', row.names=FALSE)
write.csv(gfc_stats$gain_table, 
          file='NAK_GFC_extract_thresholded_gaintable.csv', row.names=FALSE)
```

To view the output format of the CSV files output by `gfcanalysis`, see the
[loss 
table](/content/2014-03-25-analyzing-forest-change-with-gfcanalysis/NAK_GFC_extract_thresholded_gaintable.csv)
and [gain 
table](/content/2014-03-25-analyzing-forest-change-with-gfcanalysis/NAK_GFC_extract_thresholded_gaintable.csv) 
for Nam Kading.

## Making simple visualizations

There is also a function in `gfcanalysis` to calculate and save a thresholded 
annual layer stack from the GFC product (useful for simple visualizations, 
etc.):

```{r, cache=TRUE, cache.lazy=FALSE}
gfc_annual_stack <- annual_stack(gfc_thresholded)
writeRaster(gfc_annual_stack, filename="NAK_GFC_extract_thresholded_annual.envi")
```

The annual stack output by `annual_stack` has one layer for each year:

```{r, cache=TRUE, cache.lazy=FALSE}
gfc_annual_stack
```

Forest change in each year is coded as:

| Cover/Change          | Code |
|:-----------------------|:----:|
| No data               |    0 |
| Forest                |    1 |
| Non-forest            |    2 |
| Forest loss           |    3 |
| Forest gain           |    4 |
| Forest loss and gain  |    5 |
| Water                 |    6 |

The `animate_annual` function can be used to save a simple visualization of the 
thresholded annual layer stack.

Note: For this example, we are using the data in the WGS84 coordinate system. 
For a real analysis or presentation, the data should be projected into UTM or 
another projection system for this.  The `utm_zone` function in the 
`gfcanalysis` package and the `projectRaster` function in the `raster` package 
could be used to automate this. Also see the `to_utm` option for the 
`extract_gfc` function (type `?extract_gfc` in R).

To make an annual animation (in WGS84) type:

```{r, cache=TRUE, cache.lazy=FALSE}
aoi$label <- "ZOI" # Label the polygon on the plot
animate_annual(aoi, gfc_annual_stack, out_dir='.', site_name='Nam Kading')
```

The animation will be saved in the directory specified by `out_dir` (in this 
example the current working directory). To view the animation, double-click the 
new ".html" file in that directory. The animation will look [something like 
this](/content/2014-03-25-analyzing-forest-change-with-gfcanalysis/gfc_animation.html).

```{r, echo=FALSE}
# Delete the thresholded ENVI files - don't need them
unlink("NAK_GFC_extract_thresholded.*")
unlink("NAK_GFC_extract_thresholded_annual.*")
```

